A geometric framework for capturing high-resolution insights into conceptual knowledge and learning in classroom-like settings

### 2023-01-30 paper/main.tex
place \namerefs in \mbox to prevent bug when split across line or page breaks

% simple command for inline comments
\newcommand{\comment}[1]{}

@@

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work \citep{HeusEtal21} to extract the latent themes from the lectures' contents
using a topic model \citep{BleiEtal03}. Briefly, topic models take as input a
collection of text documents and learn a set of ``topics'' (i.e., latent
themes) from their contents. Once fit, a topic model can be used to transform
arbitrary (potentially new) documents into sets of ``topic proportions,''
describing the weighted blend of learned topics reflected in their texts. We
parsed automatically generated transcripts of the two lectures into
overlapping sliding windows, which we treated as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \textit{\nameref{subsec:topic-modeling}}).
Transforming these windows with our model resulted in a number-of-windows by
number-of-topics (15) topic-proportions matrix, denoting the unique mixture
of broad themes from both lectures reflected in each window's content.
Intuitively, each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional
space (whose axes are topics learned by the model). Within this space, each lecture's series of
topic vectors (i.e., corresponding to sliding windows parsed from its
transcript) forms a \textit{trajectory} that captures how its conceptual
content unfolds over time (Fig.~\ref{fig:sliding-windows}A). We resampled
these trajectories to a resolution of 1 topic vector for each second of video.

@@

@@

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \textit{\nameref{subsec:traces}}).
Essentially, the estimated knowledge at the coordinate is given by the weighted
average proportion of quiz questions the participant answered correctly, where
the weights reflect how much each question is ``about'' the content at $x$.
When we apply this approach to estimate the participant's knowledge about the
content presented in each moment of each lecture, we can obtain a detailed
timecourse describing how much ``knowledge'' the participant has about any part
of the lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also
apply this approach separately for the questions from each quiz the
participants took throughout the experiment. From just 13 questions per quiz,
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1106 samples across the two
lectures).

@@

    \caption{\textbf{Estimating moment-by-moment knowledge acquisition.}
    \textbf{A. Moment-by-moment knowledge about the \textit{Four Fundamental
    Forces}}. Each trace displays the weighted proportion of correctly answered
    questions about the content reflected in each moment of the lecture (see
    \textit{\nameref{subsec:traces}}), using responses from one
    quiz (color). The traces are averaged across participants. \textbf{B.
    Average estimated knowledge about the \textbf{Four Fundamental Forces}.}
    Each bar displays the across-timepoint average knowledge, estimated using
    the responses to one quiz's questions. \textbf{C. Moment-by-moment
    knowledge about the \textit{Birth of Stars}}. The panel is in the same
    format as Panel A, but here the knowledge estimates are for the
    moment-by-moment content of the \textit{Birth of Stars} lecture. \textbf{D.
    Average estimated knowledge about the \textit{Birth of Stars}.} The panel
    is in the same format as Panel B, but here the knowledge estimates are for
    the content of the \textit{Birth of Stars} lecture. All panels: error
    ribbons and error bars denote 95\% confidence intervals, estimated across
    participants.}

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see
    \textit{\nameref{subsec:knowledge-maps}}). The topic trajectories of each
    lecture and the coordinates of each question are indicated by dotted lines
    and dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps are
    in the same general format as the knowledge maps in Panel A, but each
    coordinate in the learning maps indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' across the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word reflected by the blend of topics
    represented at the locations of the stars in the maps. The words' colors
    indicate how much each word is weighted on average across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (blue) and
    \textit{Birth of Stars} (green) videos, respectively.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\textit{\nameref{subsec:knowledge-maps}}). Next, we sampled
points evenly from a $100 \times 100$ grid of coordinates that evenly tiled a
rectangle enclosing the 2D projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10K sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.

@@

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing lecture 1, the second between
lectures 1 and 2, and the third after viewing lecture 2 (Fig.~\ref{fig:experiment}).
The questions appearing on each quiz, for each participant, were randomly
chosen from the full set of 39, with the constraints that (a) each quiz contain 5
questions about lecture 1, 5 questions about lecture 2, and 3 questions, and
(b) each question appear exactly once. The order of questions on each quiz,
and the order of answer options for each question, were also randomized. Our
experimental protocol was approved by the Committee for the Protection of
Human Subjects at Dartmouth College.  We used the experiment to develop
and test our computational framework for estimating knowledge and learning.

@@

\subsubsection*{Constructing text embeddings of multiple lectures and questions}\label{subsec:topic-modeling}

>>>>

% simple command for inline comments
\newcommand{\comment}[1]{}
% italicize section names in \nameref and place in \mbox to prevent bug when 
% text is split across lines or pages
\NewCommandCopy{\oldnameref}{\nameref}
\renewcommand{\nameref}[1]{\mbox{\textit{\oldnameref{#1}}}}

@@

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work \citep{HeusEtal21} to extract the latent themes from the lectures' contents
using a topic model \citep{BleiEtal03}. Briefly, topic models take as input a
collection of text documents and learn a set of ``topics'' (i.e., latent
themes) from their contents. Once fit, a topic model can be used to transform
arbitrary (potentially new) documents into sets of ``topic proportions,''
describing the weighted blend of learned topics reflected in their texts. We
parsed automatically generated transcripts of the two lectures into
overlapping sliding windows, which we treated as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming these windows with our model resulted in a number-of-windows by
number-of-topics (15) topic-proportions matrix, denoting the unique mixture
of broad themes from both lectures reflected in each window's content.
Intuitively, each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional
space (whose axes are topics learned by the model). Within this space, each lecture's series of
topic vectors (i.e., corresponding to sliding windows parsed from its
transcript) forms a \textit{trajectory} that captures how its conceptual
content unfolds over time (Fig.~\ref{fig:sliding-windows}A). We resampled
these trajectories to a resolution of 1 topic vector for each second of video.

@@

Next, wondered whether we could use this same model to capture the

@@

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \nameref{subsec:traces}).
Essentially, the estimated knowledge at the coordinate is given by the weighted
average proportion of quiz questions the participant answered correctly, where
the weights reflect how much each question is ``about'' the content at $x$.
When we apply this approach to estimate the participant's knowledge about the
content presented in each moment of each lecture, we can obtain a detailed
timecourse describing how much ``knowledge'' the participant has about any part
of the lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also
apply this approach separately for the questions from each quiz the
participants took throughout the experiment. From just 13 questions per quiz,
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1106 samples across the two
lectures).

@@

    \caption{\textbf{Estimating moment-by-moment knowledge acquisition.}
    \textbf{A. Moment-by-moment knowledge about the \textit{Four Fundamental
    Forces}}. Each trace displays the weighted proportion of correctly answered
    questions about the content reflected in each moment of the lecture (see
    \nameref{subsec:traces}), using responses from one
    quiz (color). The traces are averaged across participants. \textbf{B.
    Average estimated knowledge about the \textbf{Four Fundamental Forces}.}
    Each bar displays the across-timepoint average knowledge, estimated using
    the responses to one quiz's questions. \textbf{C. Moment-by-moment
    knowledge about the \textit{Birth of Stars}}. The panel is in the same
    format as Panel A, but here the knowledge estimates are for the
    moment-by-moment content of the \textit{Birth of Stars} lecture. \textbf{D.
    Average estimated knowledge about the \textit{Birth of Stars}.} The panel
    is in the same format as Panel B, but here the knowledge estimates are for
    the content of the \textit{Birth of Stars} lecture. All panels: error
    ribbons and error bars denote 95\% confidence intervals, estimated across
    participants.}

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see
    \nameref{subsec:knowledge-maps}). The topic trajectories of each
    lecture and the coordinates of each question are indicated by dotted lines
    and dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps are
    in the same general format as the knowledge maps in Panel A, but each
    coordinate in the learning maps indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' across the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word reflected by the blend of topics
    represented at the locations of the stars in the maps. The words' colors
    indicate how much each word is weighted on average across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (blue) and
    \textit{Birth of Stars} (green) videos, respectively.}

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\nameref{subsec:knowledge-maps}). Next, we sampled
points evenly from a $100 \times 100$ grid of coordinates that evenly tiled a
rectangle enclosing the 2D projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10K sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.

@@

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing lecture 1, the second between
lectures 1 and 2, and the third after viewing lecture 2 (Fig.~\ref{fig:experiment}).
The questions appearing on each quiz, for each participant, were randomly
chosen from the full set of 39, with the constraints that (a) each quiz contain 5
questions about lecture 1, 5 questions about lecture 2, and 3 questions about general physics knowledge, and
(b) each question appear exactly once for each participant. The order of questions on each quiz,
and the order of answer options for each question, were also randomized. Our
experimental protocol was approved by the Committee for the Protection of
Human Subjects at Dartmouth College.  We used the experiment to develop
and test our computational framework for estimating knowledge and learning.

@@

\subsubsection*{Constructing text embeddings of multiple lectures and questions}\label{subsec:topic-modeling}
Following \citet{HeusEtal21}
